---
title: "VISA stat analysis: van_mic"
author: CSL
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Main comments
- Focused on columns with at least 10% ones (due to restrictions on singular data matrix)
- From nonsynonimous data:
  - only interaction of WP_000871607.1 (main effect significant 0.25941) and WP_001060462.1 (main effect not significant: 0.25742) was barely not significant (pvalue=0.054, negative coefficient: -0.34493)
  - There was no evidence that interaction effects were different by background (CC)
  - WP_000871607.1 (pvalue=0.00937, 0.20862) and WP_001788197.1 (pvalue=0.0057, 0.26263) are significantly associated with van_mic (positive coefficient)
  - growth (fitness) is not a significant covariate in the model
- From all data:
  - WP_001106057.1 (pvalue=0.0112) is significantly associated with van_mic (negative coefficient, -0.190066)
  - no interaction was significant. Closer to significance: `WP_001060462.1*WP_001106057.1` with pvalue=0.085 (positive coefficient)
  - There was no evidence that interaction effects were different by background (CC)
  - Significant interaction between WP_001788197.1 (main effect not significant -0.06457) and CC (main effect not significant -0.02670) (pvalue=0.022 with positive coefficient(WP_001788197.1:ccNRS70)=0.34003


# Using log(MIC) as response

## Nonsynonimous data: `nonsyn_var_matrix.csv`

Reading the data (starting with nonsyn data as it already has the phenotypes):
```{r}
dat = read.table("data/nonsyn_var_matrix.csv", sep=",", header=TRUE)
str(dat)
summary(dat)
```
Removing columns of all zeroes:
```{r}
dat2 = dat[, colSums(dat != 0, na.rm = TRUE) > 0]
str(dat2)
```

Removing columns of all zeroes and just one 1:
```{r}
dat3 = dat2[,c(TRUE,colSums(dat2[,2:ncol(dat2)], na.rm=TRUE) > 1)]
str(dat3)
summary(dat3)
```

Removing columns of less than 10 ones (~10%):
```{r}
dat4 = dat2[,c(TRUE,colSums(dat2[,2:ncol(dat2)], na.rm=TRUE) > 9)]
str(dat4)
summary(dat4)
```

One strange thing: I thought the columns are presence/absence, hence binary 0/1, but `WP_000918667.1` and `WP_001060462.1` have a 2:
```{r}
dat4[dat4$WP_000918667.1 == 2,]
dat4[dat4$WP_001060462.1 == 2,]

dat3[dat3$WP_001060462.1 == 2,]
dat3[dat3$WP_000918667.1 == 2,]
```
We will convert the 2 to 1:
```{r}
dat4$WP_000918667.1[dat4$WP_000918667.1 == 2] = 1
dat4$WP_001060462.1[dat4$WP_001060462.1 == 2] = 1

dat3$WP_001060462.1[dat3$WP_001060462.1 == 2] = 1
dat3$WP_000918667.1[dat3$WP_000918667.1 == 2] = 1
```

We need to remove the NA rows:
```{r}
## >10% ones
dat10.c = dat4[complete.cases(dat4),]
dat10.c = within(dat10.c, Strain<-factor(Strain))
str(dat10.c)

## >1 ones
dat1.c = dat3[complete.cases(dat3),]
dat1.c = within(dat1.c, Strain<-factor(Strain))
str(dat1.c)
```


## Statistical analyses

We will fit a linear mixed effects model using only the interactions that appeared in the plots (instead of the 15 possible interactions on 6 variables). We will group the strains by `Strain`.

Summary of interactions:

- WP_000871607.1 and WP_000918667.1: not significant (pvalue=0.63)
- WP_000871607.1 and WP_00101713.1: not significant (pvalue=0.13)
- WP_001788197.1 and WP_001017131.1: not significant (pvalue=0.23)
- WP_001788197.1 and WP_000149064.1: not significant (pvalue=0.23)
- WP_000871607.1 and WP_001060462.1: barely not significant (pvalue=0.054)
- WP_001060462.1 and WP_001788197.1: not significant (pvalue=0.3)
- WP_001017131.1 and WP_001060462.1: not significant (pvalue=0.96)

First, we fit the model without interactions:
```{r}
library(stringr)
dat10.c$cc = "NRS384"
dat10.c$cc[str_detect(dat10.c$Strain,"NRS70")] = "NRS70"
dat10.c = within(dat10.c, cc <- factor(cc))

library(lme4)
##library(lmerTest)

mod1 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod1)
```

Then, we fit the models with interactions one by one and compare to the model without interactions with ANOVA. We fit two different interaction models: one with scalar random effects (interactions do not depend on CC), and one with vector random effects (interactions depend on CC). 
```{r}
mod2 = lmer(log(van_mic) ~ WP_000871607.1*WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod2)

anova(mod1,mod2)
```
Interaction `WP_000871607.1*WP_000918667.1` not significant.


```{r}
mod3 = lmer(log(van_mic) ~ WP_000871607.1*WP_001017131.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod3)

anova(mod1,mod3)

mod3.2 = lmer(log(van_mic) ~ WP_000871607.1*WP_001017131.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_000149064.1 + (1 + WP_000871607.1*WP_001017131.1| cc), dat10.c)
summary(mod3.2)

anova(mod3,mod3.2)
anova(mod1,mod3.2)
```
Interaction `WP_000871607.1*WP_001017131.1` not significant (careful: singular solution on `mod3`).

```{r}
mod4 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1*WP_001017131.1 + WP_001060462.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod4)

anova(mod1,mod4)

mod4.2 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1*WP_001017131.1 + WP_001060462.1 + WP_000149064.1 + (1+WP_001788197.1*WP_001017131.1 | cc), dat10.c)
summary(mod4.2)

anova(mod4,mod4.2)
anova(mod1,mod4.2)
```
Interaction `WP_001788197.1*WP_001017131.1` not significant (careful: singular solution on `mod4`).

```{r}
mod5 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1*WP_000149064.1 + WP_001060462.1 + WP_001017131.1 + (1 | cc), dat10.c)
summary(mod5)

anova(mod1,mod5)
```
Interaction `WP_001788197.1*WP_000149064.1` not significant (careful: singular solution on `mod5`).

```{r}
mod6 = lmer(log(van_mic) ~ WP_000871607.1*WP_001060462.1 + WP_000918667.1 + WP_001788197.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod6)

anova(mod1,mod6)
```
Not significant interaction `WP_000871607.1*WP_001060462.1`, but pvalue=0.06.

```{r}
mod7 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1*WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod7)

anova(mod1,mod7)
```
Not significant interaction `WP_001788197.1*WP_001060462.1`.

```{r}
mod8 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1*WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod8)

anova(mod1,mod8)
```
Not significant interaction `WP_001060462.1*WP_001017131.1`.

### Standard linear model (without random effects)

Let's test if the triple interaction (two SNPs and CC) is significant on a standard linear model:
```{r}
m1 = lm(log(van_mic) ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + cc, dat10.c)
summary(m1)

## from mod6, the only interaction close to significance:
m6 = lm(log(van_mic) ~ r + WP_000871607.1*WP_001060462.1*cc + WP_000918667.1 + WP_001788197.1 + WP_001017131.1 + WP_000149064.1, dat10.c)
summary(m6)

anova(m1,m6)

## without fitness as covariate
m2 = lm(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + cc, dat10.c)
summary(m2)

anova(m2,m1)
```
Not significant either: so, no evidence of different interaction effects by cc.



## All data: `all_var_matrix.csv`

Reading the data:
```{r}
dat = read.table("data/all_var_matrix.csv", sep=",", header=TRUE)
dat0 = read.table("data/nonsyn_var_matrix.csv", sep=",", header=TRUE)

dat$r = dat0$r
dat$van_mic = dat0$van_mic
dat$dap_mic = dat0$dap_mic

str(dat)
summary(dat)
```

Removing columns of all zeroes:
```{r}
dat2 = dat[, colSums(dat != 0, na.rm = TRUE) > 0]
str(dat2)
```

Removing columns of all zeroes and just one 1:
```{r}
dat3 = dat2[,c(TRUE,colSums(dat2[,2:ncol(dat2)], na.rm=TRUE) > 1)]
str(dat3)
summary(dat3)
```

Removing columns of less than 10 ones (~10%). We have the exact same as in non-synonimous except for `WP_001106057.1`.
```{r}
dat4 = dat2[,c(TRUE,colSums(dat2[,2:ncol(dat2)], na.rm=TRUE) > 9)]
str(dat4)
summary(dat4)
```

One strange thing: I thought the columns are presence/absence, hence binary 0/1, but `WP_000918667.1, WP_001060462.1, WP_001017131.1,  WP_000149064.1` have a 2:
```{r}
dat4[dat4$WP_000918667.1 == 2,]
dat4[dat4$WP_001060462.1 == 2,]
dat4[dat4$WP_001017131.1 == 2,]
dat4[dat4$WP_000149064.1 == 2,]
```

We will convert the 2 to 1:
```{r}
dat4$WP_000918667.1[dat4$WP_000918667.1 == 2] = 1
dat4$WP_001060462.1[dat4$WP_001060462.1 == 2] = 1
dat4$WP_001017131.1[dat4$WP_001017131.1 == 2] = 1
dat4$WP_000149064.1[dat4$WP_000149064.1 == 2] = 1
```

We need to remove the NA rows:
```{r}
## >10% ones
dat10.c = dat4[complete.cases(dat4),]
dat10.c = within(dat10.c, Strain<-factor(Strain))
str(dat10.c)
```


## Statistical analyses

We will fit a linear mixed effects model using only the interactions that appeared in the plots (instead of the 15 possible interactions on 6 variables). We will group the strains by `Strain`.

Summary of interactions (all not significant):

- WP_000871607.1 and WP_001106057.1
- WP_000918667.1 and WP_001106057.1
- WP_001788197.1 and WP_001106057.1
- WP_001060462.1 and WP_001106057.1 (pvalue=0.085)
- WP_001017131.1 and WP_001106057.1
- WP_000149064.1 and WP_001106057.1

First, we fit the model without interactions:
```{r}
dat10.c$cc = "NRS384"
dat10.c$cc[str_detect(dat10.c$Strain,"NRS70")] = "NRS70"
dat10.c = within(dat10.c, cc <- factor(cc))

library(lme4)
##library(lmerTest)

mod1 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1 | cc), dat10.c)
summary(mod1)
```

Then, we fit the models with interactions one by one and compare to the model without interactions with ANOVA. We fit two different interaction models: one with scalar random effects (interactions do not depend on CC), and one with vector random effects (interactions depend on CC). 
```{r}
mod2 = lmer(log(van_mic) ~ WP_000871607.1*WP_001106057.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod2)

anova(mod1,mod2)
```
Interaction `WP_000871607.1*WP_001106057.1` not significant.


```{r}
mod3 = lmer(log(van_mic) ~ WP_000871607.1 + WP_001017131.1 + WP_000918667.1*WP_001106057.1 + WP_001788197.1 + WP_001060462.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod3)

anova(mod1,mod3)
```
Interaction `WP_000918667.1*WP_001106057.1` not significant (careful: singular solution on `mod3`).

```{r}
mod4 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1*WP_001106057.1 + WP_001017131.1 + WP_001060462.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod4)

anova(mod1,mod4)
```
Interaction `WP_001788197.1*WP_001106057.1` not significant (careful: singular solution on `mod4`).

```{r}
mod5 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_000149064.1 + WP_001060462.1*WP_001106057.1 + WP_001017131.1 + (1 | cc), dat10.c)
summary(mod5)

anova(mod1,mod5)
```
Interaction `WP_001060462.1*WP_001106057.1` not significant (careful: singular solution on `mod5`). Pvalue=0.085

```{r}
mod6 = lmer(log(van_mic) ~ WP_000871607.1 + WP_001060462.1 + WP_000918667.1 + WP_001788197.1 + WP_001017131.1*WP_001106057.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod6)

anova(mod1,mod6)
```
Not significant interaction `WP_001017131.1*WP_001106057.1` (careful: singular solution)

```{r}
mod7 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1*WP_001106057.1 + (1 | cc), dat10.c)
summary(mod7)

anova(mod1,mod7)
```
Not significant interaction `WP_000149064.1*WP_001106057.1` (careful: singular solution).

## Testing if main effects differ by CC

```{r}
mod1 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1 | cc), dat10.c)

mod1.1 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1 + WP_000871607.1 | cc), dat10.c)

anova(mod1,mod1.1)

mod1.2 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1 + WP_001788197.1 | cc), dat10.c)

anova(mod1,mod1.2)

mod1.3 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1 + WP_001106057.1 | cc), dat10.c)

anova(mod1,mod1.3)
```
No significance changes in main effects due to random effects from CC.

### Using standard linear model:
```{r}
mod1 = lm(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + cc, dat10.c)
summary(mod1)

mod1.1 = lm(log(van_mic) ~ WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + WP_000871607.1*cc, dat10.c)

anova(mod1,mod1.1)

mod1.2 = lm(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + WP_001788197.1*cc, dat10.c)
summary(mod1.2)
anova(mod1,mod1.2)

mod1.3 = lm(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1*cc, dat10.c)
summary(mod1.3)

anova(mod1,mod1.3)
```