---
title: "VISA stat analysis: dap_mic"
author: CSL
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Main comments
- Focused on columns with at least 10% ones (due to restrictions on singular data matrix)
- From nonsynonimous data:
  - only interaction of WP_000871607.1 and WP_000918667.1 was barely not significant (pvalue=0.08, negative coefficient)
  - There was no evidence that interaction effects were different by background (CC)
  - only growth rate was close to significant on main effects (pvalue=0.056)
- From all data:
  - WP_001788197.1 (pvalue=0.01, positive coefficient), WP_001017131.1 (pvalue=0.02, positive coefficient) are significantly associated with dap_mic 
  - no interaction was significant
  - There was no evidence that interaction effects were different by background (CC)


# Nonsynonimous data: `nonsyn_var_matrix.csv`

Reading the data (starting with nonsyn data as it already has the phenotypes):
```{r}
dat = read.table("data/nonsyn_var_matrix.csv", sep=",", header=TRUE)
str(dat)
summary(dat)
```
Removing columns of all zeroes:
```{r}
dat2 = dat[, colSums(dat != 0, na.rm = TRUE) > 0]
str(dat2)
```

Removing columns of all zeroes and just one 1:
```{r}
dat3 = dat2[,c(TRUE,colSums(dat2[,2:ncol(dat2)], na.rm=TRUE) > 1)]
str(dat3)
summary(dat3)
```

Removing columns of less than 10 ones (~10%):
```{r}
dat4 = dat2[,c(TRUE,colSums(dat2[,2:ncol(dat2)], na.rm=TRUE) > 9)]
str(dat4)
summary(dat4)
```

One strange thing: I thought the columns are presence/absence, hence binary 0/1, but `WP_000918667.1` and `WP_001060462.1` have a 2:
```{r}
dat4[dat4$WP_000918667.1 == 2,]
dat4[dat4$WP_001060462.1 == 2,]

dat3[dat3$WP_001060462.1 == 2,]
dat3[dat3$WP_000918667.1 == 2,]
```
We will convert the 2 to 1:
```{r}
dat4$WP_000918667.1[dat4$WP_000918667.1 == 2] = 1
dat4$WP_001060462.1[dat4$WP_001060462.1 == 2] = 1

dat3$WP_001060462.1[dat3$WP_001060462.1 == 2] = 1
dat3$WP_000918667.1[dat3$WP_000918667.1 == 2] = 1
```

We need to remove the NA rows:
```{r}
## >10% ones
dat10.c = dat4[complete.cases(dat4),]
dat10.c = within(dat10.c, Strain<-factor(Strain))
str(dat10.c)

## >1 ones
dat1.c = dat3[complete.cases(dat3),]
dat1.c = within(dat1.c, Strain<-factor(Strain))
str(dat1.c)
```

# Data Visualization plots

Now, we plot the interaction plots. We want to see if the slopes of different groups are different which could indicate potential interaction effects.

We are choosing columns one at a time to serve as the X-axis. For example, the first column `WP_000871607.1` in the X-axis, and the other columns as grouping variable for the first plots.

We are focusing on only the columns that have at least 10% ones first.

Some potential interactions (based on the plots):

- NRS70: WP_000871607.1 and WP_000918667.1
- NRS70: WP_000871607.1 and WP_001788197.1
- NRS70: WP_000871607.1 and WP_001017131.1
- NRS70: WP_000871607.1 and WP_000149064.1
- NRS70: WP_000918667.1 and WP_001788197.1
- NRS70: WP_001788197.1 and WP_001017131.1
- NRS70: WP_001788197.1 and WP_000149064.1
- NRS384: WP_000871607.1 and WP_001060462.1
- NRS384: WP_001017131.1 and WP_001060462.1


```{r, echo=FALSE}
library(gridExtra)
library(grid)
library(ggplot2)
library(stringr)

## main plotting function
interactionPlot = function(df,i,j){
  p <- ggplot(df,aes(x=df[,i],y=dap_mic, color=as.factor(df[,j])))+
          geom_jitter(alpha=0.3, width=0.05, height=0.05)+
          geom_smooth(method = "lm") +
          xlim(c(-0.1,1.1)) +
          theme(
              plot.title = element_text(hjust=0.5, size=rel(1.8)),
              axis.title.x = element_text(size=rel(1.8)),
              axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
              axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              panel.background = element_blank(),
              axis.line = element_line(colour = "grey")##,
              )
  return(p)
}

## NRS70
strain70 = "NRS70"
dat70 = dat10.c[str_detect(dat10.c$Strain,strain70),]
## NRS384
strain384 = "NRS384"
dat384 = dat10.c[str_detect(dat10.c$Strain,strain384),]

for(i in 2:(ncol(dat10.c)-3)){
  for(j in 2:(ncol(dat10.c)-3)){
    if(i != j){
      p = interactionPlot(dat70,i,j)
      p2 <- p + xlab(names(dat70)[i])+
          labs(color=names(dat70)[j]) +
          ggtitle(strain70)
      print(p2)
    
      q = interactionPlot(dat384,i,j)
      q2 <- q + xlab(names(dat384)[i])+
          labs(color=names(dat384)[j]) +
          ggtitle(strain384)
      print(q2)  
    }
  }  
}

```


# Statistical analyses

We will fit a linear mixed effects model using only the interactions that appeared in the plots (instead of the 15 possible interactions on 6 variables). We will group the strains by `Strain`.

Summary of interactions (all not significant):

- WP_000871607.1 and WP_000918667.1 (pvalue=0.08)
- WP_000871607.1 and WP_001788197.1
- WP_000871607.1 and WP_001017131.1
- WP_000871607.1 and WP_000149064.1
- WP_000918667.1 and WP_001788197.1
- WP_001788197.1 and WP_001017131.1
- WP_001788197.1 and WP_000149064.1
- WP_000871607.1 and WP_001060462.1
- WP_001017131.1 and WP_001060462.1

First, we fit the model without interactions:
```{r}
dat10.c$cc = "NRS384"
dat10.c$cc[str_detect(dat10.c$Strain,"NRS70")] = "NRS70"
dat10.c = within(dat10.c, cc <- factor(cc))

library(lme4)
##library(lmerTest)

mod1 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod1)
```

Then, we fit the models with interactions one by one and compare to the model without interactions with ANOVA. We fit two different interaction models: one with scalar random effects (interactions do not depend on CC), and one with vector random effects (interactions depend on CC). 
```{r}
mod2 = lmer(dap_mic ~ r + WP_000871607.1*WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod2)

anova(mod1,mod2)
```
Interaction `WP_000871607.1*WP_000918667.1` not significant (but close pvalue=0.08)


```{r}
mod3 = lmer(dap_mic ~ r + WP_000871607.1*WP_001017131.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod3)

anova(mod1,mod3)

mod3.2 = lmer(dap_mic ~ r + WP_000871607.1*WP_001017131.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_000149064.1 + (1 + WP_000871607.1*WP_001017131.1| cc), dat10.c)
summary(mod3.2)

anova(mod3,mod3.2)
anova(mod1,mod3.2)
```
Interaction `WP_000871607.1*WP_001017131.1` not significant (careful: singular solution on `mod3`).

```{r}
mod4 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1*WP_001017131.1 + WP_001060462.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod4)

anova(mod1,mod4)

mod4.2 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1*WP_001017131.1 + WP_001060462.1 + WP_000149064.1 + (1+WP_001788197.1*WP_001017131.1 | cc), dat10.c)
summary(mod4.2)

anova(mod4,mod4.2)
anova(mod1,mod4.2)
```
Interaction `WP_001788197.1*WP_001017131.1` not significant (careful: singular solution on `mod4`).

```{r}
mod5 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1*WP_000149064.1 + WP_001060462.1 + WP_001017131.1 + (1 | cc), dat10.c)
summary(mod5)

anova(mod1,mod5)
```
Interaction `WP_001788197.1*WP_000149064.1` not significant (careful: singular solution on `mod5`).

```{r}
mod6 = lmer(dap_mic ~ r + WP_000871607.1*WP_001060462.1 + WP_000918667.1 + WP_001788197.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod6)

anova(mod1,mod6)
```
Not significant interaction `WP_000871607.1*WP_001060462.1`, but pvalue=0.08.

```{r}
mod7 = lmer(dap_mic ~ r + WP_000871607.1*WP_001788197.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod7)

anova(mod1,mod7)
```
Not significant interaction `WP_000871607.1*WP_001788197.1`.

```{r}
mod8 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1*WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod8)

anova(mod1,mod8)
```
Not significant interaction `WP_001060462.1*WP_001017131.1`.

```{r}
mod9 = lmer(dap_mic ~ r + WP_000871607.1*WP_000149064.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + (1 | cc), dat10.c)
summary(mod9)

anova(mod1,mod9)
```
Not significant interaction `WP_000871607.1*WP_000149064.1`.

```{r}
mod10 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000149064.1 + WP_000918667.1*WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + (1 | cc), dat10.c)
summary(mod10)

anova(mod1,mod10)
```
Not significant interaction `WP_000871607.1*WP_000149064.1`.

## Standard linear model (without random effects)

Let's test if the triple interaction (two SNPs and CC) is significant on a standard linear model:
```{r}
m1 = lm(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + cc, dat10.c)
summary(m1)

## from mod6, the only interaction close to significance:
m6 = lm(dap_mic ~ r + WP_000871607.1*WP_000918667.1*cc + WP_001060462.1 + WP_001788197.1 + WP_001017131.1 + WP_000149064.1, dat10.c)
summary(m6)

anova(m1,m6)
```
Not significant either: so, no evidence of different interaction effects by cc.


# All data: `all_var_matrix.csv`

Reading the data:
```{r}
dat = read.table("data/all_var_matrix.csv", sep=",", header=TRUE)
dat0 = read.table("data/nonsyn_var_matrix.csv", sep=",", header=TRUE)

dat$r = dat0$r
dat$van_mic = dat0$van_mic
dat$dap_mic = dat0$dap_mic

str(dat)
summary(dat)
```
Removing columns of all zeroes:
```{r}
dat2 = dat[, colSums(dat != 0, na.rm = TRUE) > 0]
str(dat2)
```

Removing columns of all zeroes and just one 1:
```{r}
dat3 = dat2[,c(TRUE,colSums(dat2[,2:ncol(dat2)], na.rm=TRUE) > 1)]
str(dat3)
summary(dat3)
```

Removing columns of less than 10 ones (~10%). We have the exact same as in non-synonimous except for `WP_001106057.1`.
```{r}
dat4 = dat2[,c(TRUE,colSums(dat2[,2:ncol(dat2)], na.rm=TRUE) > 9)]
str(dat4)
summary(dat4)
```

One strange thing: I thought the columns are presence/absence, hence binary 0/1, but `WP_000918667.1, WP_001060462.1, WP_001017131.1,  WP_000149064.1` have a 2:
```{r}
dat4[dat4$WP_000918667.1 == 2,]
dat4[dat4$WP_001060462.1 == 2,]
dat4[dat4$WP_001017131.1 == 2,]
dat4[dat4$WP_000149064.1 == 2,]
```
We will convert the 2 to 1:
```{r}
dat4$WP_000918667.1[dat4$WP_000918667.1 == 2] = 1
dat4$WP_001060462.1[dat4$WP_001060462.1 == 2] = 1
dat4$WP_001017131.1[dat4$WP_001017131.1 == 2] = 1
dat4$WP_000149064.1[dat4$WP_000149064.1 == 2] = 1
```

We need to remove the NA rows:
```{r}
## >10% ones
dat10.c = dat4[complete.cases(dat4),]
dat10.c = within(dat10.c, Strain<-factor(Strain))
str(dat10.c)
```

# Data Visualization plots

Now, we plot the interaction plots. We want to see if the slopes of different groups are different which could indicate potential interaction effects.

We are choosing columns one at a time to serve as the X-axis. For example, the first column `WP_000871607.1` in the X-axis, and the other columns as grouping variable for the first plots.

We are focusing on only the columns that have at least 10% ones first, and we only focus on the extra variable that we did not have on non-synonimous data: `WP_001106057.1`

Potential interactions (based on plots):

- NRS384: WP_000871607.1 and WP_001106057.1
- NRS384: WP_000918667.1 and WP_001106057.1
- NRS384: WP_001788197.1 and WP_001106057.1
- NRS384: WP_001060462.1 and WP_001106057.1
- NRS384: WP_001017131.1 and WP_001106057.1

```{r, echo=FALSE}
library(gridExtra)
library(grid)
library(ggplot2)
library(stringr)

## main plotting function
interactionPlot = function(df,i,j){
  p <- ggplot(df,aes(x=df[,i],y=dap_mic, color=as.factor(df[,j])))+
          geom_jitter(alpha=0.3, width=0.05, height=0.05)+
          geom_smooth(method = "lm") +
          xlim(c(-0.1,1.1)) +
          theme(
              plot.title = element_text(hjust=0.5, size=rel(1.8)),
              axis.title.x = element_text(size=rel(1.8)),
              axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
              axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              panel.background = element_blank(),
              axis.line = element_line(colour = "grey")##,
              )
  return(p)
}

## NRS70
strain70 = "NRS70"
dat70 = dat10.c[str_detect(dat10.c$Strain,strain70),]
## NRS384
strain384 = "384"
dat384 = dat10.c[str_detect(dat10.c$Strain,strain384),]

for(j in 2:7){
      p = interactionPlot(dat70,8,j)
      p2 <- p + xlab(names(dat70)[8])+
          labs(color=names(dat70)[j]) +
          ggtitle(strain70)
      print(p2)
    
      q = interactionPlot(dat384,8,j)
      q2 <- q + xlab(names(dat384)[8])+
          labs(color=names(dat384)[j]) +
          ggtitle(strain384)
      print(q2)  
}
```


# Statistical analyses

We will fit a linear mixed effects model using only the interactions that appeared in the plots (instead of the 15 possible interactions on 6 variables). We will group the strains by `Strain`.

Summary of interactions (all not significant):

- WP_000871607.1 and WP_001106057.1
- WP_000918667.1 and WP_001106057.1
- WP_001788197.1 and WP_001106057.1
- WP_001060462.1 and WP_001106057.1
- WP_001017131.1 and WP_001106057.1

First, we fit the model without interactions:
```{r}
dat10.c$cc = "NRS384"
dat10.c$cc[str_detect(dat10.c$Strain,"NRS70")] = "NRS70"
dat10.c = within(dat10.c, cc <- factor(cc))

library(lme4)
##library(lmerTest)

mod1 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1 | cc), dat10.c)
summary(mod1)
```

Then, we fit the models with interactions one by one and compare to the model without interactions with ANOVA. We fit two different interaction models: one with scalar random effects (interactions do not depend on CC), and one with vector random effects (interactions depend on CC). 
```{r}
mod2 = lmer(dap_mic ~ r + WP_000871607.1*WP_001106057.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod2)

anova(mod1,mod2)
```
Interaction `WP_000871607.1*WP_001106057.1` not significant.


```{r}
mod3 = lmer(dap_mic ~ r + WP_000871607.1 + WP_001017131.1 + WP_000918667.1*WP_001106057.1 + WP_001788197.1 + WP_001060462.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod3)

anova(mod1,mod3)
```
Interaction `WP_000918667.1*WP_001106057.1` not significant (careful: singular solution on `mod3`).

```{r}
mod4 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1*WP_001106057.1 + WP_001017131.1 + WP_001060462.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod4)

anova(mod1,mod4)
```
Interaction `WP_001788197.1*WP_001106057.1` not significant (careful: singular solution on `mod4`).

```{r}
mod5 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_000149064.1 + WP_001060462.1*WP_001106057.1 + WP_001017131.1 + (1 | cc), dat10.c)
summary(mod5)

anova(mod1,mod5)
```
Interaction `WP_001060462.1*WP_001106057.1` not significant (careful: singular solution on `mod5`). Pvalue=0.1

```{r}
mod6 = lmer(dap_mic ~ r + WP_000871607.1 + WP_001060462.1 + WP_000918667.1 + WP_001788197.1 + WP_001017131.1*WP_001106057.1 + WP_000149064.1 + (1 | cc), dat10.c)
summary(mod6)

anova(mod1,mod6)
```
Not significant interaction `WP_001017131.1*WP_001106057.1` (careful: singular solution)

## Standard linear model

```{r}
m1 = lm(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + cc, dat10.c)
summary(m1)
```

## Testing if main effects differ by CC

```{r}
mod1 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1 | cc), dat10.c)

mod1.2 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1+WP_001788197.1 | cc), dat10.c)

anova(mod1,mod1.2)

mod1.3 = lmer(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1+WP_001017131.1 | cc), dat10.c)

anova(mod1,mod1.3)
```
No significant differences in main effects by random effects on CC.

Now with a standard linear model:
```{r}
mod1 = lm(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + cc, dat10.c)

mod1.2 = lm(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + WP_001788197.1*cc, dat10.c)

anova(mod1,mod1.2)

mod1.3 = lm(dap_mic ~ r + WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + WP_001017131.1*cc, dat10.c)

anova(mod1,mod1.3)
```
No significant interactions on CC either.