---
title: "VISA stat analysis summary"
author: CSL
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(lme4)
library(ggplot2)
```


# Methods

To study potential pairwise interactions between SNPs affecting the MIC of vancomycin, we fit linear mixed-effects models with presence/absence of SNPs as fixed-effects predictors (main and interaction effects) and background (NRS70, NRS384) as scalar random effect. 
In addition, we tested mixed-effects models with vector random effects (affecting intercept and slope of certain interactions), but these models were not stable enough for convergence (results not shown).
Given the sparsity of the data matrix, we restricted the predictors to common SNPs present in at least 10% of the strains. We transformed the response to logarithmic scale.

To investigate differences in the effect sizes of SNPs per background, we fit standard linear models with interaction terms for SNP presence and background. In these models, background is included as a fixed-effects predictor because the mixed-effects version (with background as vector random effect affecting the main and interaction effects of the SNPs) was not computationally stable.
 
Finally, we fitted the same models for a different response: growth rate as proxy of fitness. Similarly to the vancomycin models, we searched for pairwise SNP interactions with mixed-effects models and SNP-background interactions with standard linear models.


# Results

```{r, echo=FALSE}
dat = read.table("data/all_var_matrix.csv", sep=",", header=TRUE)
dat0 = read.table("data/nonsyn_var_matrix.csv", sep=",", header=TRUE)

dat$r = dat0$r
dat$van_mic = dat0$van_mic
dat$dap_mic = dat0$dap_mic

dat2 = dat[, colSums(dat != 0, na.rm = TRUE) > 0]
dat3 = dat2[,c(TRUE,colSums(dat2[,2:ncol(dat2)], na.rm=TRUE) > 1)]
dat4 = dat2[,c(TRUE,colSums(dat2[,2:ncol(dat2)], na.rm=TRUE) > 9)]

dat4$WP_000918667.1[dat4$WP_000918667.1 == 2] = 1
dat4$WP_001060462.1[dat4$WP_001060462.1 == 2] = 1
dat4$WP_001017131.1[dat4$WP_001017131.1 == 2] = 1
dat4$WP_000149064.1[dat4$WP_000149064.1 == 2] = 1

dat10.c = dat4[complete.cases(dat4),]
dat10.c = within(dat10.c, Strain<-factor(Strain))

dat10.c$cc = "NRS384"
dat10.c$cc[str_detect(dat10.c$Strain,"NRS70")] = "NRS70"
dat10.c = within(dat10.c, cc <- factor(cc))
```

## Interaction effects between SNPs and background

```{r, echo=FALSE}
## standard linear model
mod1 = lm(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + cc, dat10.c)

## interactions with cc
mod1.2 = lm(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + WP_001788197.1*cc, dat10.c)
a = anova(mod1,mod1.2)
pv = round(a$`Pr(>F)`[2],4)

intterm = attributes(terms(mod1.2))$term.labels[length(attributes(terms(mod1.2))$term.labels)]

### significant terms
df = summary(mod1.2)$coefficients[summary(mod1.2)$coefficients[,4]<0.05,]

str = "The significant predictors are "
for(i in 2:nrow(df)){
  w = rownames(df)[i]
  est = round(df[i,1],4)
  p = round(df[i,4],3)
  str = paste0(str, w, " (with estimated effect ", est, " and pvalue ", p, ")")
  if(i<nrow(df)){
    str = paste0(str,", ")
  }
}

## interaction plot
df2 = summary(mod1.2)$coefficients
ss = strsplit(intterm,':')
intterms = unique(c(rownames(df2)[str_detect(rownames(df2),"Intercept")], rownames(df2)[str_detect(rownames(df2),ss[[1]][1])], rownames(df2)[str_detect(rownames(df2),ss[[1]][2])]))
df3 = df2[intterms,]

## line for NRS70:
intercept1 = df3["(Intercept)",1] + df3["ccNRS70",1]
slope1 = df3["WP_001788197.1",1] + df3["WP_001788197.1:ccNRS70",1]

## line for NRS384
intercept2 = df3["(Intercept)",1]
slope2 = df3["WP_001788197.1",1]

newdf1 = data.frame(x=c(0,1),y=c(intercept1, intercept1+slope1))
newdf2 = data.frame(x=c(0,1),y=c(intercept2, intercept2+slope2))

p <- ggplot(dat10.c,aes(x=WP_001788197.1,y=log(van_mic)))+
          geom_jitter(aes(color=cc), alpha=0.3, width=0.05, height=0.05)+
          #geom_smooth(method = "lm") +
          #geom_abline(intercept=intercept1, slope=slope1, linetype="dashed", color="blue") +
          #geom_abline(intercept=intercept2, slope=slope2, linetype="dashed", color="red") +
          geom_point(data=newdf1,aes(x=x,y=y),color="deepskyblue4", size=2.5)+
          geom_line(data=newdf1,aes(x=x,y=y),color="deepskyblue4", size=1.5)+
          geom_point(data=newdf2,aes(x=x,y=y),color="firebrick3", size=2.5)+
          geom_line(data=newdf2,aes(x=x,y=y),color="firebrick3", size=1.5)+
          scale_x_continuous(breaks=c(0,1))+
          theme(
              legend.title = element_blank(),
              plot.title = element_text(hjust=0.5, size=rel(1.8)),
              axis.title.x = element_text(size=rel(1.8)),
              axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
              axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              panel.background = element_blank(),
              axis.line = element_line(colour = "grey")##,
              )
```

In the standard linear model, we included the predictors of `r attributes(terms(mod1))$term.labels`. The only significant interaction with background (`cc`) is `r intterm` with p-value `r pv`.
`r str`. Growth was not included as a confounder as it was not significant.

**Figure 1:** Effect sizes of WP_001788197.1 on the logarithm of MIC of vancomycin by background (NRS70, blue and NRS384, red). The difference in slope illustrates the interaction effect between WP_001788197.1 and background (significant with pvalue=`r pv`). The effect of WP_001788197.1 is `r newdf1$y[2]-newdf2$y[2]` higher in NRS70 than in NRS384. 

```{r, echo=FALSE}
p
```


## Pairwise interaction effects between SNPs

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## WP_001106057.1 not included due to singularity issues:
mod1 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
mod6 = lmer(log(van_mic) ~ WP_000871607.1*WP_001060462.1 + WP_000918667.1 + WP_001788197.1 + WP_001017131.1 + WP_000149064.1 + (1 | cc), dat10.c)
a2 = anova(mod1,mod6)
pv2 = round(a2$`Pr(>Chisq)`[2],4)
intterm2 = attributes(terms(mod6))$term.labels[length(attributes(terms(mod6))$term.labels)]

mod2 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1 | cc), dat10.c)
mod5 = lmer(log(van_mic) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_000149064.1 + WP_001060462.1*WP_001106057.1 + WP_001017131.1 + (1 | cc), dat10.c)
a3 = anova(mod2,mod5)
pv3 = round(a3$`Pr(>Chisq)`[2],4)
intterm3 = attributes(terms(mod5))$term.labels[length(attributes(terms(mod5))$term.labels)]

## interaction plot
df2 = summary(mod6)$coefficients
ss = strsplit(intterm2,':')
intterms = unique(c(rownames(df2)[str_detect(rownames(df2),"Intercept")], rownames(df2)[str_detect(rownames(df2),ss[[1]][1])], rownames(df2)[str_detect(rownames(df2),ss[[1]][2])]))
df3 = df2[intterms,]

## line for WP_000871607.1=0:
intercept1 = df3["(Intercept)",1]
slope1 = df3["WP_001060462.1",1]

## line for WP_000871607.1=1:
intercept2 = df3["(Intercept)",1] + df3["WP_000871607.1",1]
slope2 = df3["WP_001060462.1",1] + df3["WP_000871607.1:WP_001060462.1",1]

newdf1 = data.frame(x=c(0,1),y=c(intercept1, intercept1+slope1))
newdf2 = data.frame(x=c(0,1),y=c(intercept2, intercept2+slope2))

p2 <- ggplot(dat10.c,aes(x=WP_001060462.1,y=log(van_mic)))+
          geom_jitter(aes(color=as.factor(WP_000871607.1)), alpha=0.3, width=0.05, height=0.05)+
          geom_point(data=newdf1,aes(x=x,y=y),color="firebrick3", size=2.5)+
          geom_line(data=newdf1,aes(x=x,y=y),color="firebrick3", size=1.5)+
          geom_point(data=newdf2,aes(x=x,y=y),color="deepskyblue4", size=2.5)+
          geom_line(data=newdf2,aes(x=x,y=y),color="deepskyblue4", size=1.5)+
          scale_x_continuous(breaks=c(0,1))+
          guides(color=guide_legend(title="WP_000871607.1"))+
          theme(
              plot.title = element_text(hjust=0.5, size=rel(1.8)),
              axis.title.x = element_text(size=rel(1.8)),
              axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
              axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              panel.background = element_blank(),
              axis.line = element_line(colour = "grey")##,
              )
```

In the mixed-effects model, only one pairwise interaction was significant: `r intterm2` with pvalue=`r pv2`. The interaction of `r intterm3` was close to significant with pvalue=`r pv3`.


**Figure 2:** Effect sizes of WP_001060462.1 on the logarithm of MIC of vancomycin by WP_000871607.1 (presence: blue and absence: red). The difference in slope illustrates the interaction effect between WP_001060462.1 and WP_000871607.1 (significant with pvalue=`r pv2`). The effect of WP_001060462.1 is `r newdf1$y[2]-newdf2$y[2]` lower when WP_000871607.1 is present. 

```{r, echo=FALSE}
p2
```

## Growth rate as response

```{r, echo=FALSE}
m1 = lm(log(r) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + cc, dat10.c)
m1.4 = lm(log(r) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1*cc + WP_001017131.1 + WP_000149064.1 + WP_001106057.1, dat10.c)

a = anova(m1,m1.4)
pv = round(a$`Pr(>F)`[2],4)

intterm = attributes(terms(m1.4))$term.labels[length(attributes(terms(m1.4))$term.labels)]

### significant terms
df = summary(m1.4)$coefficients[summary(m1.4)$coefficients[,4]<0.05,]

str = "The significant predictors are "
for(i in 2:nrow(df)){
  w = rownames(df)[i]
  est = round(df[i,1],4)
  p = round(df[i,4],3)
  str = paste0(str, w, " (with estimated effect ", est, " and pvalue ", p, ")")
  if(i<nrow(df)){
    str = paste0(str,", ")
  }
}

## interaction plot
df2 = summary(m1.4)$coefficients
ss = strsplit(intterm,':')
intterms = unique(c(rownames(df2)[str_detect(rownames(df2),"Intercept")], rownames(df2)[str_detect(rownames(df2),ss[[1]][1])], rownames(df2)[str_detect(rownames(df2),ss[[1]][2])]))
df3 = df2[intterms,]

## line for NRS70:
intercept1 = df3["(Intercept)",1] + df3["ccNRS70",1]
slope1 = df3["WP_001060462.1",1] + df3["WP_001060462.1:ccNRS70",1]

## line for NRS384
intercept2 = df3["(Intercept)",1]
slope2 = df3["WP_001060462.1",1]

newdf1 = data.frame(x=c(0,1),y=c(intercept1, intercept1+slope1))
newdf2 = data.frame(x=c(0,1),y=c(intercept2, intercept2+slope2))

p3 <- ggplot(dat10.c,aes(x=WP_001060462.1,y=log(r)))+
          geom_jitter(aes(color=cc), alpha=0.3, width=0.05, height=0.05)+
          geom_point(data=newdf1,aes(x=x,y=y),color="deepskyblue4", size=2.5)+
          geom_line(data=newdf1,aes(x=x,y=y),color="deepskyblue4", size=1.5)+
          geom_point(data=newdf2,aes(x=x,y=y),color="firebrick3", size=2.5)+
          geom_line(data=newdf2,aes(x=x,y=y),color="firebrick3", size=1.5)+
          scale_x_continuous(breaks=c(0,1))+
          theme(
              legend.title = element_blank(),
              plot.title = element_text(hjust=0.5, size=rel(1.8)),
              axis.title.x = element_text(size=rel(1.8)),
              axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
              axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              panel.background = element_blank(),
              axis.line = element_line(colour = "grey")##,
              )
```

In the standard linear model, we included the predictors of `r attributes(terms(m1))$term.labels`. The only significant interaction with background (`cc`) is `r intterm` with p-value `r pv`. No predictors (aside from the interaction term) were significant.

**Figure 3:** Effect sizes of WP_001060462.1 on the logarithm of the growth rate by background (NRS70, blue and NRS384, red). The difference in slope illustrates the interaction effect between WP_001060462.1 and background (significant with pvalue=`r pv`). The effect of WP_001060462.1 is `r newdf2$y[2]-newdf1$y[2]` lower in NRS70 than in NRS384. 

```{r, echo=FALSE}
p3
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## WP_001106057.1 not included due to singularity issues:
mod1 = lmer(log(r) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1 + WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1 | cc), dat10.c)
mod13 = lmer(log(r) ~ WP_000871607.1 + WP_000918667.1 + WP_001788197.1*WP_001060462.1 + WP_001017131.1 + WP_000149064.1 + WP_001106057.1 + (1 | cc), dat10.c)
a2 = anova(mod1,mod13)
pv2 = round(a2$`Pr(>Chisq)`[2],4)
intterm2 = attributes(terms(mod13))$term.labels[length(attributes(terms(mod13))$term.labels)]


## interaction plot
df2 = summary(mod13)$coefficients
ss = strsplit(intterm2,':')
intterms = unique(c(rownames(df2)[str_detect(rownames(df2),"Intercept")], rownames(df2)[str_detect(rownames(df2),ss[[1]][1])], rownames(df2)[str_detect(rownames(df2),ss[[1]][2])]))
df3 = df2[intterms,]

## line for WP_001788197.1=0:
intercept1 = df3["(Intercept)",1]
slope1 = df3["WP_001060462.1",1]

## line for WP_001788197.1=1:
intercept2 = df3["(Intercept)",1] + df3["WP_001788197.1",1]
slope2 = df3["WP_001060462.1",1] + df3["WP_001788197.1:WP_001060462.1",1]

newdf1 = data.frame(x=c(0,1),y=c(intercept1, intercept1+slope1))
newdf2 = data.frame(x=c(0,1),y=c(intercept2, intercept2+slope2))

p4 <- ggplot(dat10.c,aes(x=WP_001060462.1,y=log(r)))+
          geom_jitter(aes(color=as.factor(WP_001788197.1)), alpha=0.3, width=0.05, height=0.05)+
          geom_point(data=newdf1,aes(x=x,y=y),color="firebrick3", size=2.5)+
          geom_line(data=newdf1,aes(x=x,y=y),color="firebrick3", size=1.5)+
          geom_point(data=newdf2,aes(x=x,y=y),color="deepskyblue4", size=2.5)+
          geom_line(data=newdf2,aes(x=x,y=y),color="deepskyblue4", size=1.5)+
          scale_x_continuous(breaks=c(0,1))+
          guides(color=guide_legend(title="WP_001788197.1"))+
          theme(
              plot.title = element_text(hjust=0.5, size=rel(1.8)),
              axis.title.x = element_text(size=rel(1.8)),
              axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
              axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
              panel.background = element_blank(),
              axis.line = element_line(colour = "grey")##,
              )
```

In the mixed-effects model, only one pairwise interaction was significant: `r intterm2` with pvalue=`r pv2`.

**Figure 4:** Effect sizes of WP_001060462.1 on the logarithm of growth rate by WP_001788197.1 (presence: blue and absence: red). The difference in slope illustrates the interaction effect between WP_001060462.1 and WP_001788197.1 (significant with pvalue=`r pv2`). The effect of WP_001060462.1 is `r newdf1$y[2]-newdf2$y[2]` lower when WP_001788197.1 is present. 

```{r, echo=FALSE}
p4
```

```{r, echo=FALSE}
newalpha = format(round(0.05/90,5), scientific=F)
```

We note that while the reported effects were considered significant under the standard significance level ($\alpha=0.05$), these effects are no longer significant after Bonferroni multiple testing correction ($\alpha=$ `r newalpha`).